---
title: "Class 8: Breast Cancer Mini Project"
author: "Lilia Jimenez (PID:A16262599)"
format: pdf
---

Before we get stuck into project work

read the data (lab 7)
```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```
>Q. How many genes are in this dataset?

```{r}
nrow(rna.data)
```


## Run PCA

```{r}
## Again we have to take the transpose of our data 
pca <- prcomp(t(rna.data), scale=TRUE)
 
## Simple un polished plot of pc1 and pc2
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```

```{r}
summary(pca)
```


```{r}
#We have 5 wt and 5 ko samples
mycols<- c(rep("blue",5), rep("red",5))
mycols

plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", col=mycols)
```

I could examibe which genes contribute the most to the first pc

```{r}
head(sort(abs(pca$rotation[,1]),decreasing = T))
```

# Analysis of Human Breast cells

```{r}

wisc.df <- read.csv("WisconsinCancer (1).csv", row.names=1)
#head(wisc.df)

```
```{r}
diagnosis<-as.factor(wisc.df$diagnosis)
```

Now I want to make sure I remove the column from my data set for analysis.

```{r}
wisc.data<-wisc.df[,-1]
#head(wisc.data)
```

Q1. How many observations are in this dataset?

```{r}
ncol(wisc.df)
```

Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```

Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.data)))

```


##Principal Component Analysis 

Here we will use 'prcomp()' on the 'wisc.data' object - the one without the diagnosis column. 

We can look at the means and sd of each column. If they are similar then we are all good to go. If not, we should use 'scale=TRUE'

```{r}
head(colMeans(wisc.data))
```

```{r}
head(apply(wisc.data, 2 ,sd))
```

```{r}
wisc.pr<- prcomp(wisc.data,scale. = TRUE)
summary(wisc.pr)
```

>Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

44.27%

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 PCs

## Plotting the PCA results

```{r}
#biplot(wisc.pr)
```

>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

It is difficult to read. Everything is just overlayed on top of eachother so I cant read most of it.


we need our own plot

```{r}
attributes(wisc.pr)
```

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=diagnosis)
```

```{r}
library(ggplot2)

pc<- as.data.frame(wisc.pr$x)

ggplot(pc)+ aes(PC1, PC2, col=diagnosis)+ geom_point()
```

#Communicating PCA results
>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean",1]
```

>Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
tbl<-summary(wisc.pr)
which(tbl$importance[3,]>0.8)[1]
```

##Hierarchial clustering

The main function for hierarchial clustering is 'hclust()' it takes a distance matrix

```{r}
d<-dist(scale(wisc.data))
wisc.hclust<-hclust(d)
plot(wisc.hclust)
```

```{r}
plot(wisc.hclust)
abline(h=18, col="red")
grps<-cutree(wisc.hclust, h=18)
table(grps)
```

Come back here. Later to see how our cluster grps correspond to M or B groups.

```{r}
#ggplot(pc)+ aes(PC1, PC2, col=diagnosis)+ geom_point()
```

##5. Combining methods

Here we will perform clustering on our PCA results rather than on original data.

In other words we will cluster using 'wisc.pr$x' - our new better variables or PCs. We can choose as many or as few PCs as we like.

```{r}
d.pc<-dist(wisc.pr$x[,1:3])

wisc.pr.hclust<- hclust(d.pc, method = "ward.D2")
plot(wisc.pr.hclust)
abline(h=80, col="red")
```

```{r}
grps<-cutree(wisc.pr.hclust, h=80)
table(grps)
```

we can use 'table()' function to make a cross-table ass as just a count table

```{r}
table(diagnosis)
```

```{r}
table(grps, diagnosis)
```

the results indicate that our cluster 1 mostly captures cancer(M)
and our cluster 2 mostly capture healthy (B) sample/individuals.

## 7. Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
head(npc)
```

and plot this up
```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```
>Q18. Which of these new patients should we prioritize for follow up based on your results?

Patients in group 2 because they are more spread out so we dont know which one truly belong in that group.