---
title: "Class 18: Pertussis mini-project"
author: "Lilia Jimenez (PID:A16262599)"
format: pdf
---

First we will examine and explore pertussis case numberss in the US as tracked by the CDC:

https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

We can use the datapasta package to scrape this data from the the website into r.

```{r}
cdc <- data.frame(year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
        cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
```

```{r}
head(cdc)
```


>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

I want to plot of cases per year with ggplot
```{r}
library(ggplot2)

ggplot(cdc)+aes(x=year,y=cases)+geom_line()+geom_vline(xintercept = 1947,col="blue")+geom_vline(xintercept = 1992, col="red")+geom_vline(xintercept = 2020, col="orange")
```
>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

Pertussis cases began to rise after this point because adap vaccines provide a smaller dosage so the immune system gets smaller effect than before.

##3. Exploring CMI-PB data

acces data from the CMI-PB project

this databas (like many modern projects) uses an API to return JSON format data.

We will use the R package 'jsonlite'

```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/subject", simplifyVector = TRUE)
head(subject)
```

>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race,subject$biological_sex)
```

```{r}
subject$year_of_birth
```

#Side note working with dates

we can use lubridate package to ease 

```{r}
library(lubridate)
```
```{r}
today()
```
```{r}
today() - ymd("2000-01-01")
```
```{r}
today()-ymd("2002-02-21")
```
```{r}
time_length( today() - ymd("2002-02-21"),  "years")
```

so what is the age of everyone on our dataset.

```{r}
subject$age<-time_length(today()-ymd(subject$year_of_birth),"years")
```

>Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

>Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```
##get more data from CMI-PB
```{r}
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = TRUE)
head(specimen)
```

>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)

meta<-inner_join(subject,specimen)
head(meta)
```

Now we can read some of the other data from CMI-PB

```{r}
ab_titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE)
head(ab_titer)
```

>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```
>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```
>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```



Our first exploratory plot:
```{r}
table(abdata$antigen)
```

```{r}
ggplot(abdata)+aes(MFI, antigen)+geom_boxplot()
```
>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

```{r}
ggplot(abdata)+aes(MFI, antigen, col=infancy_vac)+geom_boxplot()
```

there are potentially some differences here but in general it is hard to tell with this whole dataset overview...

```{r}
table(abdata$dataset)
```

lets focus in on just the 2021_dataset.


```{r}
abdata.21<-filter(abdata, dataset=="2021_dataset")
table(abdata.21$dataset)
```

```{r}
pt.21<-filter(abdata.21, isotype == "IgG",  antigen == "PT") 
```

plot

```{r}
ggplot(pt.21)+aes(x=planned_day_relative_to_boost, y=MFI_normalised, col=infancy_vac)+geom_point()
```

```{r}
ggplot(pt.21)+aes(x=planned_day_relative_to_boost, y=MFI_normalised, col=infancy_vac,group=subject_id)+geom_point()+geom_line()
```






